---
title: "00 — Get Livneh + LOCA2"
format: html
---

*This Quarto doc uses two sources of climate data:*

-   *Historical: A revised version of the* Livneh et al. data. This data has been pre-processed and is included in this repo.
-   *Future: A CMIP6 dataset downscaled via LOCA2. This data is downloaded via a [UCSD server](https://loca.ucsd.edu/loca-version-2-for-north-america-ca-jan-2023/).*

*Both sources are cropped and masked to a user-defined park boundary, then written to NetCDF files for the following, downstream QMDs.*

**User steps**

1.  Customize the `User Config` code block

2.  Verify the visual boundary plot under the `Boundary Check` code block

**Notes**

-   Running this notebook will likely take several minutes. Based on testing, runtime can vary (perhaps from external server response time). Tests with `park_code <- "CANY"` took my computer 5-10 minutes to render.

-   Outputs are saved in `data/<park_code>/processed/...` (and optionally `data/<park_code>/raw/...` ).

-   If you are running this notebook on an NPS network, you may run into issues downloading the LOCA2 data. If this happens, try running on a non-NPS network.

## User Config

```{r user-config}
# ── Park identifier (CUSTOMIZE) ───────────────────────────────────────────────
# a 4-letter NPS unit code (e.g. "CANY", "JOTR", "YOSE")
park_code <- "CANY"

# ── Climate region (CUSTOMIZE) ────────────────────────────────────────────────
# see map: https://loca.ucsd.edu/wp-content/uploads/2023/08/plot_reg.R-768x481.png
# options: "cent", "e_n_cent", "n_east", "n_west", "s_east", "s_west", "south",
#          "w_n_cent", "west", "west105"
climate_region <- "s_west"

# ── Keep raw downloads (CUSTOMIZE) ───────────────────────────────────────────
# TRUE  = raw LOCA2 files are kept in data/raw/ after processing and
#         reused on subsequent runs (skips re-downloads).
# FALSE = raw files are deleted after processing to save disk space
keep_raw_downloads <- TRUE
```

## Setup

```{r setup}
#| message: false
library(terra)
library(sf)
library(ggplot2)
library(httr)

NPS_BOUNDARY_SHP <- "data/nps_boundaries/nps_boundary.shp"
LIVNEH_DATA_DIR  <- file.path("data", "livneh")

# shared raw dir (regional LOCA2 downloads are reused across parks)
raw_dir      <- file.path("data", "raw")
loca2_dl_dir <- file.path(raw_dir, "loca2", climate_region)
# per-park output dirs
out_dir    <- file.path("data", park_code, "processed")
livneh_out <- file.path(out_dir, "livneh")
loca2_out  <- file.path(out_dir, "loca2")

for (d in c(livneh_out, loca2_dl_dir, loca2_out)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

# constants (leave these unchanged)
LIVNEH_YEARS    <- 1950:2018
LOCA2_YEARS     <- 1950:2065
LOCA2_SCENARIOS <- c("ssp245", "ssp585")
LOCA2_MODELS <- c(
  "ACCESS-CM2", "AWI-CM-1-1-MR", "BCC-CSM2-MR", "CNRM-CM6-1",
  "CNRM-ESM2-1", "EC-Earth3", "EC-Earth3-Veg", "FGOALS-g3",
  "GFDL-CM4", "GFDL-ESM4", "HadGEM3-GC31-LL", "INM-CM4-8",
  "INM-CM5-0", "IPSL-CM6A-LR", "KACE-1-0-G", "MIROC6",
  "MPI-ESM1-2-HR", "MRI-ESM2-0", "NorESM2-LM", "NorESM2-MM"
)

# one available ensemble member per model
LOCA2_ENSEMBLES <- c(
  "ACCESS-CM2"      = "r4i1p1f1", "AWI-CM-1-1-MR" = "r1i1p1f1",
  "BCC-CSM2-MR"     = "r1i1p1f1", "CNRM-CM6-1"    = "r1i1p1f2",
  "CNRM-ESM2-1"     = "r1i1p1f2", "EC-Earth3"     = "r1i1p1f1",
  "EC-Earth3-Veg"   = "r1i1p1f1", "FGOALS-g3"     = "r1i1p1f1",
  "GFDL-CM4"        = "r1i1p1f1", "GFDL-ESM4"     = "r1i1p1f1",
  "HadGEM3-GC31-LL" = "r1i1p1f3", "INM-CM4-8"     = "r1i1p1f1",
  "INM-CM5-0"       = "r1i1p1f1", "IPSL-CM6A-LR"  = "r1i1p1f1",
  "KACE-1-0-G"      = "r1i1p1f1", "MIROC6"        = "r1i1p1f1",
  "MPI-ESM1-2-HR"   = "r1i1p1f1", "MRI-ESM2-0"    = "r1i1p1f1",
  "NorESM2-LM"      = "r1i1p1f1", "NorESM2-MM"    = "r1i1p1f1"
)

LOCA2_BASE_URL <- "https://cirrus.ucsd.edu/~pierce/LOCA2/CONUS_regions_split"
LOCA2_RES <- "0p0625deg"

LOCA2_DATE_RANGES <- c("2015-2044", "2045-2074")

PR_TO_MM_YR <- 86400 * 365.25 # kg m-2 s-1 -> mm yr-1
K_TO_C <- -273.15 # K -> degC

terraOptions(progress = 0)  # suppress terra progress bars
```

## Load Boundary

```{r boundary}
#| message: false
nps_all <- st_read(NPS_BOUNDARY_SHP, quiet = TRUE)

if (!park_code %in% nps_all$UNIT_CODE) {
  stop(
    "'", park_code, "' is not a valid NPS unit code. ",
    "Check the UNIT_CODE field in ", NPS_BOUNDARY_SHP, "."
  )
}

boundary_wgs84 <- nps_all[nps_all$UNIT_CODE == park_code, ] |>
  st_transform("EPSG:4326")
boundary_vect <- vect(boundary_wgs84) |> buffer(width = 500)  # 500 m buffer to capture edge plots

park_name <- unique(boundary_wgs84$UNIT_NAME)
message("Loaded boundary for: ", park_name, " (", park_code, ")")
```

## Boundary Check

Quick visual sanity check of the boundary we provided.

```{r boundary-check}
#| fig-width: 5
#| fig-height: 4
ggplot() +
  geom_sf(data = boundary_wgs84, fill = "grey85", color = "black", linewidth = 0.7) +
  labs(
    title    = paste("Boundary check –", park_name),
    subtitle = paste0("NPS unit code: ", park_code)
  ) +
  theme_minimal()
```

## Historical Livneh Dataset

This data is pre-built (see `build-livneh-regions.R`) into one file per LOCA2 climate region. Here, we read the region matching `climate_region`, then crop and mask it to the park boundary.

```{r livneh-historical}
#| message: false
# LOCA2 also serves a "west105" folder (a separate resolution of the same
# west region); Livneh only has one file for it, named "west".
livneh_region <- if (climate_region == "west105") "west" else climate_region

livneh_pr_r     <- rast(file.path(LIVNEH_DATA_DIR, paste0(livneh_region, "_pr.nc")))
livneh_tasmax_r <- rast(file.path(LIVNEH_DATA_DIR, paste0(livneh_region, "_tasmax.nc")))

bv <- project(boundary_vect, crs(livneh_pr_r))
livneh_pr     <- crop(mask(livneh_pr_r, bv, touches = TRUE), bv, snap = "out") # mm/yr
livneh_tasmax <- crop(mask(livneh_tasmax_r, bv, touches = TRUE), bv, snap = "out") # degC

suppressWarnings(writeRaster(livneh_pr, file.path(livneh_out, "livneh_pr.nc"), filetype = "netCDF", overwrite = TRUE))
suppressWarnings(writeRaster(livneh_tasmax, file.path(livneh_out, "livneh_tasmax.nc"), filetype = "netCDF", overwrite = TRUE))

message("Wrote livneh_pr.nc and livneh_tasmax.nc")
```

## LOCA2

### Download

For each model, scenario, and date-range chunk, we grab the appropriate NetCDF from the UCSD server:

```{r loca2-helpers}
list_loca2_dir <- function(url) {
  resp <- GET(url, timeout(60))
  if (status_code(resp) == 404L) return(character())
  stop_for_status(resp)
  html <- content(resp, as = "text", encoding = "UTF-8")
  m <- gregexpr('href="([^"?/][^"]*)"', html, perl = TRUE)
  hrefs <- regmatches(html, m)[[1]]
  gsub('href="|"', "", hrefs) |> trimws() |> sub(pattern = "/$", replacement = "")
}

# return the best matching filename for a model/scenario/variable/date_range.
find_loca2_file <- function(model, scenario, variable, date_range) {
  ensemble <- LOCA2_ENSEMBLES[[model]]
  url <- paste(LOCA2_BASE_URL, model, climate_region, LOCA2_RES,
               ensemble, scenario, variable, "", sep = "/")
  files <- list_loca2_dir(url)
  cands <- files[grepl(date_range, files, fixed = TRUE) & grepl("\\.nc$", files)]
  # prefer yearly
  yearly <- cands[grepl("\\.yearly\\.", cands) & !grepl("all_mean", cands)]
  if (length(yearly)) return(list(url = url, file = sort(yearly)[1]))
  # fall back to monthly
  monthly <- cands[grepl("\\.monthly\\.", cands) & !grepl("clim", cands)]
  if (length(monthly)) return(list(url = url, file = sort(monthly)[1]))
  NULL
}
```

```{r loca2-download}
#| message: false
for (model in LOCA2_MODELS) {
  for (scenario in LOCA2_SCENARIOS) {
    for (date_range in LOCA2_DATE_RANGES) {
      for (variable in c("pr", "tasmax")) {
        dest_dir <- file.path(loca2_dl_dir, model, scenario, variable)
        # skip HTTP listing if a file matching this date_range is already cached
        if (length(list.files(dest_dir, pattern = date_range)) > 0) {
          message("[skip-cached] ", model, "/", scenario, "/", variable, "/", date_range)
          next
        }
        found <- find_loca2_file(model, scenario, variable, date_range)
        if (is.null(found)) {
          message("[skip] ", model, "/", scenario, "/", variable, "/", date_range, " — not found")
          next
        }
        dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
        dest_file <- file.path(dest_dir, found$file)
        dl_url    <- paste0(found$url, found$file)
        message("[dl] ", model, "/", scenario, "/", variable, "/", found$file)
        download.file(dl_url, dest_file, mode = "wb", quiet = TRUE)
      }
    }
  }
}
```

### Load + Convert LOCA2 Variables

This helper loads all downloaded files for a given model/scenario/variable. For `tasmax`, which is downloaded as monthly data, it computes the annual mean. For both variables, it applies unit conversions.

```{r loca2-annual-tasmax}
# for a given model/scenario, load all downloaded tasmax files
# and return a list(arr, years) where arr is (time, lat, lon) in degC.
load_loca2_var <- function(model, scenario, variable) {
  var_dir <- file.path(loca2_dl_dir, model, scenario, variable)
  files   <- sort(list.files(var_dir, pattern = "\\.nc$", full.names = TRUE))
  if (!length(files)) return(NULL)

  chunks <- lapply(files, function(f) {
    r <- rotate(rast(f))  # LOCA2 uses 0-360 lon so we rotate to -180/+180
    n_layers <- nlyr(r)
    is_monthly <- n_layers > (LOCA2_YEARS[length(LOCA2_YEARS)] - LOCA2_YEARS[1] + 1) * 1.5

    if (is_monthly) {
      yrs <- as.integer(format(time(r), "%Y"))
      unique_yrs <- sort(unique(yrs))
      annual_layers <- lapply(unique_yrs, function(y) mean(r[[which(yrs == y)]]))
      r <- rast(annual_layers)
      time(r) <- as.Date(paste0(unique_yrs, "-07-01"))
    }

    # unit conversion
    if (variable == "pr") {
      r <- r * PR_TO_MM_YR
    } else {
      r <- r + K_TO_C
    }
    list(r = r, years = as.integer(format(time(r), "%Y")))
  })

  # concatenate chunks along time
  r_all    <- rast(lapply(chunks, `[[`, "r"))
  yrs_all  <- unlist(lapply(chunks, `[[`, "years"))
  list(r = r_all, years = yrs_all)
}
```

### Standardize + Crop LOCA2

For each model and scenario, we crop and mask to the boundary and write two NetCDFs per model–scenario pair (e.g. `ACCESS-CM2_ssp245_pr.nc` and `ACCESS-CM2_ssp245_tasmax.nc`), one per variable.

```{r loca2-standardize}
#| message: false
process_var <- function(vd) {
  if (is.null(vd)) return(NULL)
  in_range <- vd$years >= min(LOCA2_YEARS) & vd$years <= max(LOCA2_YEARS)
  bv       <- project(boundary_vect, crs(vd$r))
  if (!relate(bv, as.polygons(ext(vd$r)), "intersects")) return(NULL)
  r        <- crop(mask(vd$r[[which(in_range)]], bv, touches = TRUE), bv, snap = "out")
  time(r)  <- as.Date(paste0(vd$years[in_range], "-07-01"))
  r
}

for (model in LOCA2_MODELS) {
  message("Processing ", model, " ...")
  for (scenario in LOCA2_SCENARIOS) {
    pr_r     <- process_var(load_loca2_var(model, scenario, "pr"))
    tasmax_r <- process_var(load_loca2_var(model, scenario, "tasmax"))

    if (is.null(pr_r) && is.null(tasmax_r)) {
      message("  [skip] ", model, "/", scenario, " — no data or no overlap with boundary")
      next
    }

    base <- file.path(loca2_out, paste0(model, "_", scenario))
    if (!is.null(pr_r))
      suppressWarnings(writeRaster(pr_r,     paste0(base, "_pr.nc"),     filetype = "netCDF", overwrite = TRUE))
    if (!is.null(tasmax_r))
      suppressWarnings(writeRaster(tasmax_r, paste0(base, "_tasmax.nc"), filetype = "netCDF", overwrite = TRUE))
    message("  -> ", base, "_{pr,tasmax}.nc")
  }
}
```

### Clean Up Raw LOCA2

```{r loca2-cleanup}
if (keep_raw_downloads) {
  message("Keeping raw LOCA2 downloads in ", loca2_dl_dir)
} else {
  unlink(loca2_dl_dir, recursive = TRUE)
  message("Removed raw LOCA2 downloads")
}
```
