---
title: "Climate Futures Model Selection"
format: html
---

*This Quarto doc takes processed PRISM + CMIP6 LOCA2 data (downloaded via `00-get-prism-loca2.qmd`). It classifies the CMIP6 model runs into climate futures (warm-wet, hot-dry, etc.) following the [Climate Futures for Planning](https://irma.nps.gov/DataStore/FileSource/Get?id=2302720&filename=2302720.html) framework. For a given park code, it produces a quadrant scatter plot and writes a CSV of the selected model members for later plotting.*

**User steps**

1.  Customize the `User Config` code block

    -   The `selected_futures` var can be selected to align with CCRP park-specific futures or simply based on user interest. CCRP park-specific futures have been defined in the park's respective [Climate Futures Summaries](https://www.nps.gov/subjects/climatechange/climatefutures.htm) report (click "Open this panel to choose your park report," then select your park of interest on the website)
    -   The `..._start` and `..._end` year vars are loosely based on the CCRP Climate Futures [methods report](https://irma.nps.gov/DataStore/FileSource/Get?id=2302720&filename=2302720.html). The default ranges should be appropriate for most cases, but feel free to change as needed

2.  If you'd like, examine the quadrant plot and model members of the selected climate futures

**Notes**

-   Outputs are saved in `data/<park_code>/climate-futures-models.csv`

## User Config

```{r user-config}
# ── Park identifier (CUSTOMIZE) ───────────────────────────────────────────────
park_code <- "CANY"

# ── Climate futures to include in the output CSV (CUSTOMIZE) ──────────────────
# Options: "warm-wet", "warm-dry", "hot-dry", "hot-wet", "central"
selected_futures <- c("warm-wet", "hot-dry")

# ── Time periods for anomaly calculation (CUSTOMIZE) ──────────────────────────
# customize these or leave them as is
# baseline/historical period 
baseline_start <- 1984
baseline_end <- 2014
# future window
midcent_start <- 2035
midcent_end <- 2065
```

## Setup

```{r setup}
#| message: false
library(terra)
library(ggplot2)

# (feel free to ignore these constants, they're set to match the processed data)
CLIMATE_DATA_DIR <- file.path("data", park_code, "processed")
OUTPUT_CSV <- file.path("data", park_code, "climate-futures-models.csv")
LOCA2_SCEN_YEARS <- list(historical = 1950:2014, ssp245 = 2015:2065, ssp585 = 2015:2065)
LOCA2_MODELS <- unique(sub("_(historical|ssp245|ssp585)_(pr|tasmax)\\.nc$", "",
                            list.files(file.path(CLIMATE_DATA_DIR, "loca2"), pattern = "\\.nc$")))
PRISM_YEARS <- 1950:2024
PRISM_PR_NC <- file.path(CLIMATE_DATA_DIR, "prism_pr.nc")
PRISM_TASMAX_NC <- file.path(CLIMATE_DATA_DIR, "prism_tasmax.nc")
```

## Helper Functions

```{r helpers}
extract_ts_mean <- function(r) {
  global(r, "mean", na.rm = TRUE)$mean
}
```

## Compute Mid-Century Anomalies

Here we compute the mid-century mean anomaly in `tasmax` and `pr` relative to the observed PRISM baseline for each of the 40 model/scenario pairs (20 models x SSP2-4.5 + SSP5-8.5).

```{r compute-anomalies}
#| message: false
# pre-compute PRISM observed baselines
baseline_idx <- which(PRISM_YEARS >= baseline_start & PRISM_YEARS <= baseline_end)
prism_baseline <- list(
  tasmax = mean(extract_ts_mean(rast(PRISM_TASMAX_NC)[[baseline_idx]])),
  pr     = mean(extract_ts_mean(rast(PRISM_PR_NC)[[baseline_idx]]))
)

# opens each model/scenario file and returns mid-century anomalies
compute_midcent_anomalies <- function(model, scenario) {
  nc_base  <- file.path(CLIMATE_DATA_DIR, "loca2", paste0(model, "_", scenario))
  scen_yrs <- LOCA2_SCEN_YEARS[[scenario]]
  fut_idx  <- which(scen_yrs >= midcent_start & scen_yrs <= midcent_end)

  fut_tas <- mean(extract_ts_mean(rast(paste0(nc_base, "_tasmax.nc"))[[fut_idx]]))
  fut_pr  <- mean(extract_ts_mean(rast(paste0(nc_base, "_pr.nc"))[[fut_idx]]))

  list(
    tas = fut_tas - prism_baseline$tasmax,
    pr  = (fut_pr - prism_baseline$pr) / prism_baseline$pr * 100
  )
}

model_grid <- expand.grid(model    = LOCA2_MODELS,
                          scenario = c("ssp245", "ssp585"),
                          stringsAsFactors = FALSE)

anomaly_df <- do.call(rbind, lapply(seq_len(nrow(model_grid)), function(i) {
  anoms <- compute_midcent_anomalies(model_grid$model[i], model_grid$scenario[i])
  data.frame(model    = model_grid$model[i],
             scenario = model_grid$scenario[i],
             tas      = anoms$tas,
             pr       = anoms$pr)
}))
```

## Classify into Climate Futures

Now we can assign the runs a climate future label based on where its temperature and precipitation anomalies fall relative to the ensemble's 25th, 50th, and 75th percentiles.

```{r classify}
q_tas <- quantile(anomaly_df$tas, c(0.25, 0.50, 0.75))
q_pr  <- quantile(anomaly_df$pr,  c(0.25, 0.50, 0.75))

classify_future <- function(tas, pr, q_tas, q_pr) {
  ifelse(
    (tas < q_tas["25%"] & pr < q_pr["50%"]) | (tas < q_tas["50%"] & pr < q_pr["25%"]), "warm-dry",
  ifelse(
    (tas < q_tas["25%"] & pr > q_pr["50%"]) | (tas < q_tas["50%"] & pr > q_pr["75%"]), "warm-wet",
  ifelse(
    (tas > q_tas["75%"] & pr < q_pr["50%"]) | (tas > q_tas["50%"] & pr < q_pr["25%"]), "hot-dry",
  ifelse(
    (tas > q_tas["75%"] & pr > q_pr["50%"]) | (tas > q_tas["50%"] & pr > q_pr["75%"]), "hot-wet",
    "central"
  ))))
}

classified_df <- transform(anomaly_df,
  climate_future = classify_future(tas, pr, q_tas, q_pr)
)

table(classified_df$climate_future)
```

## Quadrant Plot

```{r quadrant-plot}
#| fig-width: 8
#| fig-height: 6
future_colors <- c(
  "hot-dry"  = "#b5381c",
  "hot-wet"  = "#2c7bb6",
  "warm-dry" = "#e88b29",
  "warm-wet" = "#1a9641",
  "central"  = "gray50"
)

med_tas <- q_tas["50%"]
med_pr  <- q_pr["50%"]

ssp245_df <- subset(classified_df, scenario == "ssp245")
ssp585_df <- subset(classified_df, scenario == "ssp585")

ggplot(classified_df, aes(x = tas, y = pr)) +
  # Quadrant background shading (anchored to ensemble medians)
  annotate("rect", xmin = -Inf,    xmax = med_tas, ymin = med_pr, ymax = Inf,
           fill = future_colors["warm-wet"], alpha = 0.10) +
  annotate("rect", xmin = -Inf,    xmax = med_tas, ymin = -Inf,   ymax = med_pr,
           fill = future_colors["warm-dry"], alpha = 0.10) +
  annotate("rect", xmin = med_tas, xmax = Inf,     ymin = med_pr, ymax = Inf,
           fill = future_colors["hot-wet"],  alpha = 0.10) +
  annotate("rect", xmin = med_tas, xmax = Inf,     ymin = -Inf,   ymax = med_pr,
           fill = future_colors["hot-dry"],  alpha = 0.10) +
  # IQR box
  annotate("rect",
           xmin = q_tas["25%"], xmax = q_tas["75%"],
           ymin = q_pr["25%"],  ymax = q_pr["75%"],
           fill = NA, color = "gray50", linetype = "dashed", linewidth = 0.6) +
  # Median lines
  geom_vline(xintercept = med_tas, color = "gray50", linetype = "dashed") +
  geom_hline(yintercept = med_pr,  color = "gray50", linetype = "dashed") +
  # SSP2-4.5: filled circles (fill = climate future color, black border)
  geom_point(data = ssp245_df,
             aes(fill = climate_future),
             shape = 21, color = "black", size = 3, stroke = 0.7) +
  # SSP5-8.5: hollow circles (outline = climate future color)
  geom_point(data = ssp585_df,
             aes(color = climate_future),
             shape = 21, fill = NA, size = 3, stroke = 0.7) +
  # Quadrant corner labels
  annotate("text", x = -Inf, y = Inf,  label = "Warm\u2013Wet",
           hjust = -0.1, vjust = 1.4, color = "gray30", fontface = "bold", size = 3.5) +
  annotate("text", x = -Inf, y = -Inf, label = "Warm\u2013Dry",
           hjust = -0.1, vjust = -0.4, color = "gray30", fontface = "bold", size = 3.5) +
  annotate("text", x = Inf,  y = Inf,  label = "Hot\u2013Wet",
           hjust = 1.1,  vjust = 1.4, color = "gray30", fontface = "bold", size = 3.5) +
  annotate("text", x = Inf,  y = -Inf, label = "Hot\u2013Dry",
           hjust = 1.1,  vjust = -0.4, color = "gray30", fontface = "bold", size = 3.5) +
  scale_fill_manual(values = future_colors, name = "Climate future") +
  scale_color_manual(values = future_colors, guide = "none") +
  scale_x_continuous(expand = expansion(mult = 0.12)) +
  scale_y_continuous(expand = expansion(mult = 0.12)) +
  labs(
    title    = paste("Climate Futures \u2013", park_code),
    subtitle = paste0(
      "Baseline: ", baseline_start, "\u2013", baseline_end,
      "   |   Mid-century: ", midcent_start, "\u2013", midcent_end,
      "   \u25cf SSP2-4.5 (filled)   \u25cb SSP5-8.5 (hollow)"
    ),
    x = "Temperature anomaly (\u00b0C)",
    y = "Precipitation anomaly (%)"
  ) +
  theme_bw()
```

## Export CSV

Finally, we filter to the selected climate futures and write the CSV.

```{r export-csv}
output_df <- subset(classified_df, climate_future %in% selected_futures)
output_df <- output_df[order(output_df$climate_future, output_df$model), ]

# Capitalise first letter of label (e.g. "warm-wet" → "Warm-wet")
cap_first <- function(x) paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
output_df$label <- cap_first(output_df$climate_future)

dir.create(dirname(OUTPUT_CSV), showWarnings = FALSE, recursive = TRUE)
write.csv(output_df[, c("label", "model", "scenario")],
          OUTPUT_CSV, row.names = FALSE)
message("Wrote ", nrow(output_df), " model/scenario pairs to: ", OUTPUT_CSV)

# Warn if any selected future has no model/scenario pairs
empty_futures <- selected_futures[
  !selected_futures %in% output_df$climate_future
]
if (length(empty_futures) > 0) {
  warning(
    "The following selected climate future(s) have 0 models: ",
    paste(empty_futures, collapse = ", "), ".\n",
    "Consider selecting a different future from: ",
    paste(setdiff(unique(classified_df$climate_future), selected_futures), collapse = ", "), "."
  )
}

knitr::kable(output_df[, c("label", "model", "scenario")], row.names = FALSE)
```
