---
title: "Plot PRISM + LOCA2"
format: html
---

*This Quarto doc takes in processed PRISM + CMIP6 LOCA2 data. For a given park code and its specific climate futures model members (via `01-get-climate-futures.qmd`), it provides a few temporal and spatial EDA plots.*

**User steps**

1.  Customize the `User Config` code block

    -   The `start` and `end` year vars should be set to the periods you are interested in. For example, if I plan on using the pipeline to forecast from 2025-2045, those could be my `future_start` and `future_end` values

2.  If you'd like, examine the temporal and spatial plots

## User Config

```{r user-config}
# ── Park identifier (CUSTOMIZE) ───────────────────────────────────────────────
park_code <- "CANY"

# ── Time ranges (CUSTOMIZE) ───────────────────────────────────────────────────
# PRISM historical data range: 1950-2024
# LOCA2 projection data range: 2015-2065
hist_start <- 1994  # start of historical plot window
hist_end <- 2024  # end of historical plot window
future_start <- 2025  # start of future plot window
future_end <- 2045  # end of future plot window
```

## Setup

```{r setup}
#| message: false
library(terra)
library(sf)
library(ggplot2)
library(tidyterra)
library(patchwork)

NPS_BOUNDARY_SHP <- "data/nps_boundaries/nps_boundary.shp"

# (feel free to ignore these constants, they're set to match the processed data)
CLIMATE_DATA_DIR    <- file.path("data", park_code, "processed")
CLIMATE_FUTURES_CSV <- file.path("data", park_code, "climate-futures-models.csv")
PRISM_PR_NC <- file.path(CLIMATE_DATA_DIR, "prism_pr.nc")
PRISM_TASMAX_NC  <- file.path(CLIMATE_DATA_DIR, "prism_tasmax.nc")
PRISM_YEARS <- 1950:2024
LOCA2_SCEN_YEARS <- list(historical = 1950:2014, ssp245 = 2015:2065, ssp585 = 2015:2065)
```

## Helper Functions

These are just some helper functions to prepare the data for plotting.

```{r helpers}
#| message: false
nps_all        <- st_read(NPS_BOUNDARY_SHP, quiet = TRUE)
boundary_wgs84 <- nps_all[nps_all$UNIT_CODE == park_code, ] |>
  st_transform("EPSG:4326")
park_name      <- unique(boundary_wgs84$UNIT_NAME)

# Returns one spatial mean per layer (data is pre-masked in 00-get-prism-loca2).
extract_ts_mean <- function(r) {
  global(r, "mean", na.rm = TRUE)$mean
}

# Collapses all layers to a single time-mean SpatRaster.
extract_spatial_mean <- function(r) {
  mean(r)
}

# Build scenario groups from CSV; order of groups follows the CSV.
.scen_df   <- read.csv(CLIMATE_FUTURES_CSV, stringsAsFactors = FALSE)
.scen_lbls <- unique(.scen_df$label)
scenarios  <- setNames(
  lapply(.scen_lbls, function(lbl) {
    df <- .scen_df[.scen_df$label == lbl, ]
    list(label = lbl, members = df[, c("model", "scenario")])
  }),
  .scen_lbls
)

# assign plot colours
.pal <- c("steelblue", "tomato", "forestgreen", "darkorange", "purple")
group_colors <- c("PRISM" = "black",
                  setNames(.pal[seq_along(scenarios)], names(scenarios)))
roll_n <- 10 # rolling average window in years (for timeseries plots)
```

## Load Data

Now we load the PRISM + LOCA2 data using the helpers defined above.

### PRISM

```{r load-prism}
#| message: false
hist_mask <- PRISM_YEARS >= hist_start & PRISM_YEARS <= hist_end

prism_ppt_r  <- rast(PRISM_PR_NC)
prism_tmax_r <- rast(PRISM_TASMAX_NC)

prism_ppt_df <- data.frame(year   = PRISM_YEARS[hist_mask],
                           value  = extract_ts_mean(prism_ppt_r[[hist_mask]]),
                           source = "PRISM")
prism_tmax_df <- data.frame(year   = PRISM_YEARS[hist_mask],
                            value  = extract_ts_mean(prism_tmax_r[[hist_mask]]),
                            source = "PRISM")

prism_ppt_rast  <- extract_spatial_mean(prism_ppt_r[[hist_mask]])
prism_tmax_rast <- extract_spatial_mean(prism_tmax_r[[hist_mask]])
```

### LOCA2

```{r load-loca2}
#| message: false
# Returns a data frame (year, value, source) averaged across all group members.
load_loca2_ts <- function(group, variable) {
  rows <- lapply(seq_len(nrow(group$members)), function(i) {
    model    <- group$members$model[i]
    scenario <- group$members$scenario[i]
    scen_yrs <- LOCA2_SCEN_YEARS[[scenario]]
    yr_mask  <- scen_yrs >= future_start & scen_yrs <= future_end
    r        <- rast(file.path(CLIMATE_DATA_DIR, "loca2",
                               paste0(model, "_", scenario, "_", variable, ".nc")))
    data.frame(year  = scen_yrs[yr_mask],
               value = extract_ts_mean(r[[which(yr_mask)]]))
  })
  agg        <- aggregate(value ~ year, data = do.call(rbind, rows), FUN = mean)
  agg$source <- group$label
  agg
}

# Returns a single-layer SpatRaster: spatial mean over future_start–future_end,
# averaged across all group members.
load_loca2_rast <- function(group, variable) {
  rasters <- lapply(seq_len(nrow(group$members)), function(i) {
    model    <- group$members$model[i]
    scenario <- group$members$scenario[i]
    scen_yrs <- LOCA2_SCEN_YEARS[[scenario]]
    yr_mask  <- scen_yrs >= future_start & scen_yrs <= future_end
    r        <- rast(file.path(CLIMATE_DATA_DIR, "loca2",
                               paste0(model, "_", scenario, "_", variable, ".nc")))
    extract_spatial_mean(r[[which(yr_mask)]])
  })
  Reduce("+", rasters) / length(rasters)
}

loca2_ppt_df  <- do.call(rbind, lapply(scenarios, load_loca2_ts, variable = "pr"))
loca2_tmax_df <- do.call(rbind, lapply(scenarios, load_loca2_ts, variable = "tasmax"))

loca2_ppt_rasts  <- lapply(scenarios, load_loca2_rast, variable = "pr")
loca2_tmax_rasts <- lapply(scenarios, load_loca2_rast, variable = "tasmax")
```

## Sanity check — boundary alignment

Quick visual sanity check of the boundary we provided.

```{r sanity-check}
#| fig-width: 5
#| fig-height: 4
ggplot() +
  geom_spatraster(data = prism_ppt_rast) +
  geom_sf(data = boundary_wgs84, fill = NA, color = "red", linewidth = 0.6) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       na.value = NA, name = "mm/yr") +
  labs(title    = paste("Boundary check –", park_name),
       subtitle = paste0("PRISM mean precipitation ", hist_start, "–", hist_end)) +
  theme_minimal()
```

## Time-Series Plots

If the sanity check looks good, we can take a look at some timeseries plots.

```{r ts-helper}
make_ts_plot <- function(prism_df, loca2_df, y_label, title) {
  plot_df        <- rbind(prism_df, loca2_df)
  plot_df$source <- factor(plot_df$source,
                            levels = c("PRISM", names(scenarios)))

  # Compute centred rolling mean per source group
  roll_df <- do.call(rbind, lapply(split(plot_df, plot_df$source), function(d) {
    d        <- d[order(d$year), ]
    d$roll   <- as.numeric(stats::filter(d$value, rep(1 / roll_n, roll_n), sides = 2))
    d
  }))

  ggplot(plot_df, aes(x = year, y = value, color = source)) +
    geom_line(data = roll_df, aes(y = roll), linewidth = 0.6,
              alpha = 0.5, na.rm = TRUE) +
    geom_line(linewidth = 0.8) +
    scale_color_manual(values = group_colors) +
    labs(title    = title,
         subtitle = paste0(roll_n, "-year rolling average (lighter line)"),
         x = "Year", y = y_label, color = NULL) +
    theme_minimal()
}
```

```{r ts-ppt}
make_ts_plot(prism_ppt_df, loca2_ppt_df,
             y_label = "Mean Annual Precipitation (mm/yr)",
             title   = paste("Annual Precipitation –", park_name))
```

```{r ts-tmax}
make_ts_plot(prism_tmax_df, loca2_tmax_df,
             y_label = "Mean Annual Max Temperature (°C)",
             title   = paste("Annual Max Temperature –", park_name))
```

## Spatial Anomaly Maps

And finally some spatial plots.

```{r spatial-helper}
# prism_rast  : single-layer SpatRaster (PRISM baseline, native resolution)
# group_rasts : named list of single-layer SpatRasters (one per scenario group)
# anom_fn     : function(future_rast, baseline_rast) -> anomaly SpatRaster
spatial_panels <- function(prism_rast, group_rasts,
                            base_palette, base_name,
                            anom_palette, anom_direction, anom_name,
                            anom_fn, title) {
  # Reproject PRISM baseline to LOCA2 grid for anomaly calculation
  prism_resampled <- project(prism_rast, group_rasts[[1]], method = "bilinear")
  anom_rasts <- lapply(group_rasts, function(gr) anom_fn(gr, prism_resampled))
  anom_lim   <- max(abs(unlist(lapply(anom_rasts, function(r) values(r, na.rm = TRUE)))),
                    na.rm = TRUE)

  cany_vect  <- vect(boundary_wgs84)
  cany_prism <- project(cany_vect, prism_rast)
  cany_loca2 <- project(cany_vect, group_rasts[[1]])

  theme_map <- theme_void() +
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5, size = 10))

  p_base <- ggplot() +
    geom_spatraster(data = prism_rast) +
    geom_spatvector(data = cany_prism, fill = NA, color = "white", linewidth = 0.4) +
    scale_fill_distiller(palette = base_palette, direction = 1,
                         na.value = NA, name = base_name) +
    labs(title = paste0("PRISM mean\n(", hist_start, "–", hist_end, ")")) +
    theme_map

  make_anom_panel <- function(r, group_label) {
    ggplot() +
      geom_spatraster(data = r) +
      geom_spatvector(data = cany_loca2, fill = NA, color = "white", linewidth = 0.4) +
      scale_fill_distiller(palette = anom_palette, direction = anom_direction,
                           limits = c(-anom_lim, anom_lim),
                           na.value = NA, name = anom_name) +
      labs(title = paste0(group_label, "\nvs PRISM (",
                          future_start, "–", future_end, ")")) +
      theme_map
  }

  anom_panels <- mapply(make_anom_panel, anom_rasts, names(group_rasts),
                        SIMPLIFY = FALSE)
  Reduce("|", c(list(p_base), anom_panels)) +
    plot_annotation(title = title,
                    theme = theme(plot.title = element_text(hjust = 0.5,
                                                             face = "bold")))
}
```

```{r spatial-ppt}
#| fig-width: 10
#| fig-height: 4
spatial_panels(
  prism_ppt_rast, loca2_ppt_rasts,
  base_palette   = "Blues",  base_name      = "mm/yr",
  anom_palette   = "BrBG",   anom_direction = 1,  anom_name = "% change",
  anom_fn        = function(fut, base) (fut - base) / base * 100,
  title          = paste("Mean Annual Precipitation –", park_name)
)
```

```{r spatial-tmax}
#| fig-width: 10
#| fig-height: 4
spatial_panels(
  prism_tmax_rast, loca2_tmax_rasts,
  base_palette   = "YlOrRd", base_name      = "°C",
  anom_palette   = "RdBu",   anom_direction = -1, anom_name = "Δ°C",
  anom_fn        = function(fut, base) fut - base,
  title          = paste("Mean Annual Max Temperature –", park_name)
)
```
